---
title: "Week2"
output:
  html_document: default
  pdf_document: default
date: "2024-03-08"
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Task 1: 
## The data is obviously not pre-processed – can you spot the problems?
Yes, there is a few questions in that data set: 
1.The dataset contains merged cells, which makes it easy for 
humans to read but difficult for machines to process efficiently.
2. Additionally, there are missing data points in the Excel file.

## How would you address the challenges identified above? Briefly summarise your thoughts.
First, I will create an additional column named 'description,'
 which will contain a brief description of each status. Then, 
 I will utilize the 'pivot_longer' method to transform 
 each column along with its corresponding rows.

 # Task 2: Clean the data

```{r, message = FALSE}
# import libs
library(readxl) 
library(zoo)
library(purrr)
library(tidyverse)

# loading data 
raw.df <- read_excel("/Users/vicmon/stat765/STATS765_UOA/Week_2/simulation_outpu1_raw.xlsx", col_name = FALSE) #full path comment when submiting 
# raw.df <- read_excel"Week_2/simulation_outpu1_raw.xlsx", col_name = FALSE)


# Create column names 
header_names <- expand.grid(x = c('small', 'medium', 'large'),
                            y = c('offices'),
                            z = c('concrete', 'curtainWall')) %>% 
  mutate(nn = paste(x, y, z, sep='_')) %>% 
  pull(nn)  


names(raw.df) <- c('weather', 'description', header_names)


## dealing with merged cells, create 'group' e.g. overall, sensible cooling etc
x1 <- as.data.frame(matrix(rep(NA, length(raw.df)), nrow = 1))
names(x1) <- names(raw.df)
df2 <- bind_rows(x1, raw.df) %>% 
  mutate(group = ifelse(is.na(description), 'overall',
                        ifelse(description=='Sensible Cooling', 'sensible_cooling',
                               ifelse(description=='Sensible Heating', 'sensible_heating', NA)))) %>% 
  fill(group, .direction = 'down') %>% 
  fill(weather, .direction = 'down') %>% 
  filter(!(description %in%c(NA, 'Sensible Cooling',
                             'Sensible Heating'))) %>% 
  pivot_longer(cols = small_offices_concrete:large_offices_curtainWall,
               names_to = 'CATEGORY',
               values_to = 'value') %>% 
  separate(CATEGORY, c('office_size', 'office', 'office_type'),
           sep='_') %>% 
  select(-office)

glimpse(raw.df)
summary(raw.df)
```

# task 3: Exploration using {ggplot2}

## filtering with Cooling condition
```{r, message = FALSE}
# cooling condition 
cool <- df2 %>% filter(group %in% c("sensible_cooling" ))
```

## plotting Calculated Design Air Flow

```{r, message = FALSE}
# plotting Calculated Design Air Flow by weather condition
desc_air_flow <- cool %>%  filter(description %in% c("Calculated Design Air Flow [m3/s]"))
desc_air_flow%>%
  filter(office_size %in% c("small", "medium","large") )%>% 
  ggplot(aes(y = weather, x = value, color= interaction(office_size, office_type))) +
  geom_jitter()

```

## plotting user design air Flow
```{r, message = FALSE}
desc_peak <- cool %>% filter( description %in% c("User Design Air Flow [m3/s]")) #%>%  filter(weather %in% c("Very Hot Humid (Honolulu, HI)")) 

desc_peak %>%
  ggplot(aes(y = weather, x = value, color= interaction(office_size, office_type))) +
  geom_jitter()
```



## plotting Thermostat Setpoint Temperature at Peak Load by weather condition
```{r}
desc_temp <- cool %>% filter(description %in% c("Thermostat Setpoint Temperature at Peak Load [C]"))
desc_temp %>%
  ggplot(aes(y = weather, x = value, color= interaction(office_size, office_type))) +
  geom_jitter()
```

## plotting Outdoor Temperature at Peak Load by weather condition
```{r, message = FALSE} 
desc_out_door <- cool %>%  filter(description %in% c("Outdoor Temperature at Peak Load [C]"))
desc_out_door %>%     ggplot(aes(y = weather, x = value, color = interaction(office_size, office_type))) +
  geom_jitter()

```

## plotting Outdoor Humidity Ratio at Peak Load by weather condition
``` {r, message = FALSE} 
desc_hum_rat <- cool %>%  filter((description %in% c("Outdoor Humidity Ratio at Peak Load [kgWater/kgAir]")))
desc_hum_rat  %>%   ggplot(aes(y = weather, x = value, color =  interaction(office_size, office_type))) +
  geom_jitter()
```

# Plotting heating condition(most of just repeat from above only changes the cooling to heating) 
```{r, message = FALSE}
heat <- df2 %>%  filter(group %in% c("sensible_heating"))
```
```{r, message = FALSE}
# plotting Calculated Design Air Flow
heat_air_flow <- heat %>%  filter(description %in% c("Calculated Design Air Flow [m3/s]")) %>%
  filter(office_size %in% c("small", "medium","large") )%>% 
  ggplot(aes(y = weather, x = value, color= interaction(office_size, office_type))) +
  geom_jitter()
```
```{r, message = FALSE}
# plotting User Design Air Flow 
heat_peak <- heat %>% filter( description %in% c("User Design Air Flow [m3/s]")) #%>%  filter(weather %in% c("Very Hot Humid (Honolulu, HI)")) 

heat_peak %>%
  ggplot(aes(y = weather, x = value, color= interaction(office_size, office_type))) +
  geom_jitter()
```


```{r, message = FALSE}
# listing Thermostat Setpoint Temperature at Peak Load by weather condition
heat_temp <- heat %>% filter(description %in% c("Thermostat Setpoint Temperature at Peak Load [C]"))
heat_temp %>%
  ggplot(aes(y = weather, x = value, color= interaction(office_size, office_type))) +
  geom_jitter()
# listing Outdoor Temperature at Peak Load by weather condition
heat_out_door <- heat %>%  filter(description %in% c("Outdoor Temperature at Peak Load [C]"))
heat_out_door %>%     ggplot(aes(y = weather, x = value, color = interaction(office_size, office_type))) +
  geom_jitter()
```
```{r, message = FALSE}
# plotting Outdoor Humidity Ratio at Peak Load by weather condition
heat_hum_rat <- heat %>%  filter((description %in% c("Outdoor Humidity Ratio at Peak Load [kgWater/kgAir]")))
heat_hum_rat  %>%   ggplot(aes(y = weather, x = value, color =  interaction(office_size, office_type))) +
  geom_jitter()

```


